Cosmological Hydrodynamic Simulations of Preferential Accretion in the SMBH of Milky Way Size Galaxies

### 2016-09-30 ms.tex
Added Kelly and Marta's references (mostly)

Supermassive black holes (SMBHs) are thought to exist in almost all massive galaxies \citep[see][for a review]{Kormendy2013}. In the canonical picture of BH growth, these black holes may become active galactic nuclei (AGN) during periods of high accretion and wane in periods of quiescence \citep{Alexander2005,Papovich2006,Volonteri2012}. The host galaxy's size, star formation rate, and other environmental effects may help to influence the growth of the black hole residing at its center; however, there are still uncertainties concerning the relationship between these SMBHs and their much larger host galaxies, as well as how they grow and evolve together \citep{Haehnelt2000,DiMatteo2005,Hopkins2006,Fu2008,Sijacki2009,Silverman2009,Mullaney2012}.

The M$-\sigma$ relation, which relates the SMBH's mass and the velocity dispersion of the host galaxy's central stellar population, gives some insight into the complex interplay between these objects \citep{Ferrarese2000}. A prominent trend appears, as SMBHs tend to scale with the velocity dispersion of the host galaxy bulge. The tightness of the relation is significant and can be seen over several orders of magnitudes in velocity dispersion and black hole mass \citep[e.g.][]{Merritt2001,Graham2011,Mcconnell2013,Kormendy2013}. Scatter exists among the low mass galaxies and a deviation may appear at the high mass end, where overmassive BHs may reside \citep{VanDenBosch2007,Moster2010,Natarajan2011}. However, scatter in less massive galaxies may imply that there are several channels of black hole growth at play in the low mass end of the relation \citep{Micic2007,Volonteri2009,Reines2013,Graham2014}. One standard explanation for the M$-\sigma$ relation lies in galaxy mergers, which build up galaxies, feed SMBHs, and assemble bulges \citep[e.g.][]{DiMatteo2005,Shen2008}. Major mergers are thought to supply gas to the central SMBH resulting in feedback which quenches star formation and affects the structure of the galaxy.

% Kelly cut: In addition to the M$-\sigma$, other relations between the SMBH and its host have been well studied, making it clear that, though these objects have many orders of magnitude between their sizes, a complex interplay still exists.

%[CITE FOR THEORY: Springel and Dimatteo; cite when we discuss why m-sigma happens]
%McConnell and Ma have the prettiest M-sigma relation, but there are two camps: Kormendy and Ferrarese & Merritt who must both be cited.
%Check with Kelly: Which Van den Bosch, Emsellem, and Prija papers she would recommend.

@@

Major mergers between massive galaxies are thought to be efficient fueling mechanisms for bright AGN. The large influx of material due to tidal torques from the merger causes bursts of star formation and helps funnel gas directly into the center where the SMBH resides \citep[e.g.][]{Richards2006,Reddy2008,Hopkins2010}. Additionally, the most massive, highest-luminosity AGN (i.e. quasars) reside in incredibly luminous infrared galaxies where star formation is abundant, signifying that major mergers may have recently occurred \citep{Treister2012}. Distorted morphologies are often characteristics of quasar hosts, and companions can also be present around quasars, both of which are evidence that strengthen the possibility of a recent merger having affected their lifetimes.

In many less massive and less luminous AGN, however, there is a clear lack of distorted morphology, close neighbors, and/or other obvious merger evidence \citep{Ryan2007,Schawinski2011,Ellison2013,Hicks2013}. It is also important to note that many of these AGN exist in spiral galaxies, which are unlikely to have been recently disturbed by major mergers \citep{Schawinski2011,Kocevski2011}. Nevertheless, some evidence suggests that disturbed galaxies may reform a disk quickly, even after a major merger, as long as it is gas-rich \citep{vanGorkom1997}. The rapid disk reformation of the galaxy in this study was previously studied by \cite{Governato2009} (See \ref{results}). More recently, \cite{Treister2012} has suggested that only the highest luminosity AGN require fueling via major mergers; $\sim$ 90 $\%$ of AGN across all redshifts are fueled by various other mechanisms, which may include minor mergers, flybys, and smooth accretion, whereby gas is directly accreted via large filaments from the ambient intergalactic medium \citep{Cox2006,Bellovary2013,Sinha2012}.

Smooth accretion, in particular, may play an important role in fueling these low mass galaxies. Halos less than 10$^{11}$ $M_{\odot}$ can accrete filaments of unshocked gas; thereafter, gas will shock heat to the virial temperature of the halo \citep{Keres2005}. Even for massive halos, unshocked gas may still penetrate shocked regions to fuel the galaxy \citep{Brooks2009,Dekel2009,Nelson2013}. In addition, SMBH feedback, the depositing of energy and angular momentum back into the gas reservoir during accretion, also affects the overarching structure of the host galaxy \citep{Governato2009a}. Secular processes, including bar formation and disk instabilities, may also be prominent forms of accretion for these SMBHs \citep{Kormendy2013}.

@@

%\subsection{High Resolution CHANGA Simulations}
%\textcolor{violet}{
We note that a lower resolution version of h258 was run using the Gasoline code \citep{Wadsley2004}. Our higher resolution h258 run has a spline force softening length of 174 pc and initial gas particle masses of 2.7 $\times$ $10^4 M_{\odot}$. Star particles are created with 30$\%$ of their parent gas particle mass, allowing a maximum initial mass of 8100 $M_{\odot}$. Halo h258 contains about 5 million DM particles inside the virial radius at z=0 and over 14 million DM, star, and gas particles total. The resolution of both force and mass in these simulations is comparable to the ``Eris'' simulation which has one of the highest resolutions for an N-body+SPH cosmological simulation of a Milky Way-mass galaxy so far produced.

@@

Supernova feedback releases $10^{51}$ ergs of thermal energy within a ``blastwave'' radius determined by the equations of \cite{Ostriker1988}. In the affected region, cooling turns off for a time relative to the expansion phase of the SN remnant also determined by the blastwave equation. SN Ia and II from \cite{Thielemann1986} and \cite{Woosley1995}, respectively, are implemented through the \cite{Raiteri1996} method, which uses the stellar lifetime calculations of the Padova group \citep{Alongi1993, Bressan1993, Bertelli1994} to describe stars with varying metallicities. Both the supernova ``blastwave'' radius and supernova (Ia and II) prescriptions are described in detail by \cite{Stinson2006}.

@@

%\textcolor{violet}{
We first identify halos using the Amiga Halo Finder which uses an overdensity criterion for a flat universe \citep{Knebe2001,Knollmann2009,Gill2004} to set the virial radius. We select the primary halo to be the most massive at z=0 in the high resolution region. The central SMBH in the primary halo has a mass of $1.3 \times 10^{7} M_{\odot}$.  
%}

In this analysis, we retrace each gas particle that was accreted by the galaxy or SMBH, following the gas back through its journey in the galaxy and recording its host halo and time of accretion \citep{Brooks2009}. The particles are then classified into types by their method of entrance into the primary halo. In particular, gas particles are classified as ``early'' if they exist in the primary halo before z $\sim$ 16. Gas that belonged to a different halo than the primary prior to accretion is classified as ``clumpy,'' entering the primary halo through mergers. All other gas is classified as ``smooth'' accretion, and is then subdivided into two categories: ``unshocked'' and ``shocked.'' Unshocked gas will usually flow into the halo via large-scale, dark matter filaments \citep{Keres2005,Bellovary2013}. It is possible for unshocked gas to be dense enough to pierce an already developed shock, allowing it to funnel into the galaxy core where it can be accreted onto the SMBH.

@@

%\textcolor{violet}{
To better understand the apparent preference for merger-accreted gas, we examine the angular momentum of the gas at the moment it enters the halo. Figure \ref{hrh258angmom} shows a cumulative distribution of the angular momentum of the gas as it enters the halo (solid lines). We further distinguish the gas that enters the SMBH (dashed lines), still considering its angular at the moment of halo entry. The gas is again distinguished by origin (clumpy, unshocked, shocked being green, blue, and red, respectively). We find that the angular momentum of gas entering which eventually reaches the SMBH is lower overall, and that the lowest angular momentum gas is comprised of both clumpy and unshocked gas. This result can be seen in Figure \ref{hrh258angmom_merger2}, which shows the cumulative distribution of the angular momentum of the incoming gas particles at the time of the most massive merger at z $\sim$ 1.2. (Colors and linestyles as in Figure \ref{hrh258angmom}.) Figure \ref{hrh258angmom_merger2} explicitly shows that the gas ending up in the SMBH enters with the lowest angular momentum. This influx of clumpy gas primarily had its origins in the secondary halo. Examining the angular momentum of all the gas entering from the secondary halo, we found that it was consistent with the angular momentum of the total clumpy gas entering the main halo. 
%}

@@

%\textcolor{violet}{
A lower resolution version of the galaxy h258 was run using the N-body code, Gasoline, like the simulations in \cite{Bellovary2013}. Despite fundamental differences in the hydrodynamic implementation and gas physics included and the absence of a dynamical friction prescription, an analysis of this low resolution h258 results in a SMBH with the same distinct preference for accreting clumpy gas. A second galaxy, h277, was also run using Gasoline; however, this galaxy was characterized by a quiescent merger history (no major mergers) and resulted in a SMBH whose accreted mass fractions mirrored the halo (as seen in \cite{Bellovary2013}). The broad consistency between the low and high resolution simulation of the same galaxy, and the similar results of a quiescent galaxy to the previous study, indicates that the large scale gravitational dynamics could be a main driver of the SMBH fueling in this case. We also stress that that while major mergers may not be the only physical mechanisms by which gas can be funneled into the centers of galaxies (the previous study being a strong example of this), mergers between galaxies clearly play an important role when considering the gas accretion of SMBHs.
%}

>>>>

Supermassive black holes (SMBHs) are thought to exist in almost all massive galaxies \citep[see][for a review]{Kormendy2013}. In the canonical picture of BH growth, these black holes may become active galactic nuclei (AGN) during periods of high accretion and wane in periods of quiescence \citep{Begelman1980,Alexander2005,Papovich2006,Volonteri2012}. The host galaxy's size, star formation rate, and other environmental effects may help to influence the growth of the black hole residing at its center; however, there are still uncertainties concerning the relationship between these SMBHs and their much larger host galaxies, as well as how they grow and evolve together \citep{Haehnelt2000,DiMatteo2005,Hopkins2006,Fu2008,Sijacki2009,Silverman2009,Micic2011,Mullaney2012}.

The M$-\sigma$ relation, which relates the SMBH's mass and the velocity dispersion of the host galaxy's central stellar population, gives some insight into the complex interplay between these objects \citep{Ferrarese2000,Gebhardt2000}. A prominent trend appears, as SMBHs tend to scale with the velocity dispersion of the host galaxy bulge. The tightness of the relation is significant and can be seen over several orders of magnitudes in velocity dispersion and black hole mass \citep[e.g.][]{Merritt2001,Tremaine2002,Graham2011,Mcconnell2013}. Scatter exists among the low mass galaxies and a deviation may appear at the high mass end, where overmassive BHs may reside \citep{VanDenBosch2007,Moster2010,Natarajan2011,Emsellem2011,Dubois2015}. However, scatter in less massive galaxies may imply that there are several channels of black hole growth at play in the low mass end of the relation \citep{Micic2007,Volonteri2009,Reines2013,Graham2014}. One standard explanation for the M$-\sigma$ relation lies in galaxy mergers, which build up galaxies, feed SMBHs, and assemble bulges \citep[e.g.][]{DiMatteo2005,Shen2008}. Major mergers are thought to supply gas to the central SMBH resulting in feedback which quenches star formation and affects the structure of the galaxy \citep{Schawinski2010}.

% Kelly cut: In addition to the M$-\sigma$, other relations between th,e SMBH and its host have been well studied, making it clear that, though these objects have many orders of magnitude between their sizes, a complex interplay still exists.

%[CITE FOR THEORY: Springel and Dimatteo; cite when we discuss why m-sigma happens]
%McConnell and Ma have the prettiest M-sigma relation, but there are two camps,: Kormendy and Ferrarese & Merritt who must both be cited.
%Check with Kelly: Which Van den Bosch, Emsellem, and Prija papers she would recommend.

@@

Major mergers between massive galaxies are thought to be efficient fueling mechanisms for bright AGN. The large influx of material due to tidal torques from the merger causes bursts of star formation and helps funnel gas directly into the center where the SMBH resides \citep[e.g.][]{Richards2006,Reddy2008,Hopkins2010}. Additionally, the most massive, highest-luminosity AGN (i.e. quasars) reside in incredibly luminous infrared galaxies where star formation is abundant, signifying that major mergers may have recently occurred \citep{Treister2012}. Distorted morphologies are often characteristics of quasar hosts, and companions can also be present around quasars, both of which are evidence that strengthen the possibility of a recent merger having affected their lifetimes \citep{Ellison2010}.

In many less massive and less luminous AGN, however, there is a clear lack of distorted morphology, close neighbors, and/or other obvious merger evidence \citep{Ryan2007,Schawinski2011,Ellison2013,Hicks2013}. It is also important to note that many of these AGN exist in spiral galaxies, which are unlikely to have been recently disturbed by major mergers \citep{Schawinski2011,Kocevski2011}. Nevertheless, some evidence suggests that disturbed galaxies may reform a disk quickly, even after a major merger, as long as it is gas-rich \citep{vanGorkom1997}. The rapid disk reformation of the galaxy in this paper was previously studied by \cite{Governato2009} (See \ref{results}). More recently, \cite{Treister2012} has suggested that only the highest luminosity AGN require fueling via major mergers; $\sim$ 90 $\%$ of AGN across all redshifts are fueled by various other mechanisms, which may include minor mergers, flybys, and smooth accretion, whereby gas is directly accreted via large filaments from the ambient intergalactic medium \citep{Cox2006,Bellovary2013,Sinha2012}.

Smooth accretion, in particular, may play an important role in fueling these low mass galaxies. Halos less than 10$^{11}$ $M_{\odot}$ can accrete filaments of unshocked gas; thereafter, gas will shock heat to the virial temperature of the halo \citep{Keres2005}. Even for massive halos, unshocked gas may still penetrate shocked regions to fuel the galaxy \citep{Brooks2009,Dekel2009,Nelson2013}. In addition, SMBH feedback, the depositing of energy and angular momentum back into the gas reservoir during accretion, also affects the overarching structure of the host galaxy \citep{Governato2009a}. Secular processes, including bar formation and disk instabilities, may also be prominent forms of accretion for these SMBHs \citep{Athanassoula2016,Kormendy2013}.

@@

%\subsection{High Resolution CHANGA Simulations}
%\textcolor{violet}{
We note that a lower resolution version of h258 was run using the Gasoline code \citep{Wadsley2004}. Our higher resolution h258 run has a spline force softening length of 174 pc and initial gas particle masses of 2.7 $\times$ $10^4 M_{\odot}$. Star particles are created with 30$\%$ of their parent gas particle mass, allowing a maximum initial mass of 8100 $M_{\odot}$. Halo h258 contains about 5 million DM particles inside the virial radius at z=0 and over 14 million DM, star, and gas particles total. The resolution of both force and mass in these simulations is comparable to the ``Eris'' simulation which has one of the highest resolutions for an N-body+SPH cosmological simulation of a Milky Way-mass galaxy so far produced \citep{Guedes2011}.

@@

Supernova feedback releases $10^{51}$ ergs of thermal energy within a ``blastwave'' radius determined by the equations of \cite{Ostriker1988}. In the affected region, cooling turns off for a time relative to the expansion phase of the SN remnant also determined by the blastwave equation. SN Ia and II rates from \cite{Thielemann1986} and \cite{Woosley1995}, respectively, are implemented through the \cite{Raiteri1996} method, which uses the stellar lifetime calculations of the Padova group \citep{Alongi1993, Bressan1993, Bertelli1994} to describe stars with varying metallicities. Both the supernova ``blastwave'' radius and supernova (Ia and II) prescriptions are described in detail by \cite{Stinson2006}.

@@

%\textcolor{violet}{
We first identify halos using the Amiga Halo Finder which uses an overdensity criterion for a flat universe \citep{Knebe2001,Knollmann2009,Gill2004} to set the virial radius in the primary halo. We select the primary halo to be the most massive at z=0 in the high resolution region. The central SMBH in the primary halo has a mass of $1.3 \times 10^{7} M_{\odot}$.  
%}

In this analysis, we retrace each gas particle that was accreted by the galaxy or SMBH, following the gas back through its journey in the galaxy and recording its host halo and time of accretion \citep{Brooks2009}. The particles are then classified into types by their method of entrance into the primary halo. In particular, gas particles are classified as ``early'' if they exist in the primary halo before z $\sim$ 16. Gas that belonged to a different halo than the primary prior to accretion is classified as ``clumpy,'' entering the primary halo through mergers. All other gas is classified as ``smooth'' accretion, and is then subdivided into two categories: ``unshocked'' and ``shocked.'' Unshocked gas will usually flow into the halo via large-scale, dark matter filaments \citep{Keres2005,Bellovary2013}. It is possible for unshocked gas to be dense enough to pierce an already developed shock, allowing it to funnel into the galaxy core where it can be accreted onto the SMBH \citep{Nelson2013}.

@@

%\textcolor{violet}{
To better understand the apparent preference for merger-accreted gas, we examine the angular momentum of the gas at the moment it enters the halo. Figure \ref{hrh258angmom} shows a cumulative distribution of the angular momentum of the gas as it enters the halo (solid lines). We further distinguish the gas that enters the SMBH (dashed lines), still considering its angular at the moment of halo entry. The gas is again distinguished by origin (clumpy, unshocked, shocked being green, blue, and red, respectively). We find that the angular momentum of gas entering which eventually reaches the SMBH is lower overall, and that the lowest angular momentum gas is comprised of both clumpy and unshocked gas. This result can be seen in Figure \ref{hrh258angmom_merger2}, which shows the cumulative distribution of the angular momentum of the incoming gas particles at the time of the most massive merger at z $\sim$ 1.2. (Colors and linestyles as in Figure \ref{hrh258angmom}.) Figure \ref{hrh258angmom_merger2} explicitly shows that the gas ending up in the SMBH enters with the lowest angular momentum. This influx of clumpy gas primarily had its origins in the secondary halo. Examining the angular momentum of all the gas entering from the secondary halo, we found that it was consistent with the angular momentum of the total clumpy gas entering the main halo. This may imply that the low angular momentum of the gas accreted onto the SMBH may have some of its angular momentum due to the physical properties of the merger \citep{Capelo2015}.

%}

@@

%\textcolor{violet}{
A lower resolution version of the galaxy h258 was run using the N-body code, Gasoline, like the simulations in \cite{Bellovary2013}. Despite fundamental differences in the hydrodynamic implementation and gas physics included and the absence of a dynamical friction prescription, an analysis of this low resolution h258 results in a SMBH with the same distinct preference for accreting clumpy gas. A second galaxy, h277, was also run using Gasoline; however, this galaxy was characterized by a quiescent merger history (no major mergers) and resulted in a SMBH whose accreted mass fractions mirrored the halo (as seen in \cite{Bellovary2013}). The broad consistency between the low and high resolution simulation of the same galaxy, and the similar results of a quiescent galaxy to the previous study, indicates that the large scale gravitational dynamics could be a main driver of the SMBH fueling in this case. We also stress that that while major mergers may not be the only physical mechanisms by which gas can be funneled into the centers of galaxies (the previous study being a strong example of this), mergers between galaxies clearly play an important role when considering the gas accretion of SMBHs.
%}{}
